Relative parabolic factorizations of Chevalley groups

### 2015-08-31 dennis-vaserstein.tex
Reordered lemmata in section 2

Let $\Phi$ be an irreducible root system of rank $\ell$.
For an integer $1 \leq s \leq \ell$ consider the following two subgroups of $\E(\Phi, R)$:
$$\EP_s(R) = \E(\Delta_s, R) \cdot \U(\Sigma_s, R),\ \EP_s(R, I) = \E(\Delta_s, R, I) \cdot \U(\Sigma_s, I).$$

Let $1\leq r,s \leq \ell$ be two distinct integers.
From Levi decomposition it follows that the following four subsets of $\E(\Phi, R, I)$ are equal:
\begin{multline*}
\U(\Phi^+, I)\cdot \U(\Phi^-, I) \cdot \E(\Delta_r, R, I) \cdot \EP_s(R, I) = \\
= \U(\Sigma_r, I)\cdot \U(\Sigma^-_r, I) \cdot \E(\Delta_r, R, I) \cdot \EP_s(R, I) = \hspace{5em} \\
\hspace{5em} = \EP_r(R, I) \cdot \E(\Delta_s, R, I) \cdot \U(\Sigma_s^-, I)\cdot \U(\Sigma_s, I) = \\
= \EP_r(R, I) \cdot \U(\Sigma^-_r\cap \Sigma^-_s, I) \cdot \EP_s(R, I).
\end{multline*}
\begin{dfn}
 Denote by $A_{rs}$ any of the above equal subsets.
 We call the decomposition of the form $\E(\Phi, R, I) = A_{rs}$ the {\it relative Dennis---Vaserstein} decomposition.
\end{dfn}

>>>>

Let $\Phi$ be an irreducible root system of rank $\ell$ and let $1\leq r,s \leq \ell$ be two distinct integers.
From Levi decomposition~\ref{rel:Levi-decomp} it follows that the following four subsets of $\E(\Phi, R, I)$ are equal:
\begin{multline*}
\U(\Phi^+, I)\cdot \U(\Phi^-, I) \cdot \E(\Delta_r, R, I) \cdot \EP_s(R, I) = \\
= \U(\Sigma_r, I)\cdot \U(\Sigma^-_r, I) \cdot \E(\Delta_r, R, I) \cdot \EP_s(R, I) = \hspace{5em} \\
\hspace{5em} = \EP_r(R, I) \cdot \E(\Delta_s, R, I) \cdot \U(\Sigma_s^-, I)\cdot \U(\Sigma_s, I) = \\
= \EP_r(R, I) \cdot \U(\Sigma^-_r\cap \Sigma^-_s, I) \cdot \EP_s(R, I).
\end{multline*}

\begin{lemma}\label{lemma:dv-normal} The set $A_{rs}$ is normalized by the subgroup $\EP_{\{r,s\}}(R) \leq \E(\Phi, R)$. \end{lemma}

\begin{dfn}
 Denote by $A_{rs}$ any of the above equal subsets.
 We call the decomposition of the form $\E(\Phi, R, I) = A_{rs}$ the {\it relative Dennis---Vaserstein} decomposition.
\end{dfn}

### 2015-08-31 paper.tex
Reordered lemmata in section 2

%Math operators
\DeclareMathOperator{\K}{K}
\DeclareMathOperator{\SK}{SK}
\DeclareMathOperator{\G}{G}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\SL}{SL}
\DeclareMathOperator{\Sp}{Sp}
\DeclareMathOperator{\St}{St}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\EP}{EP}
\DeclareMathOperator{\Par}{P}
\DeclareMathOperator{\B}{B}
\DeclareMathOperator{\Hh}{H}
\DeclareMathOperator{\U}{U}
\DeclareMathOperator{\Z}{Z}
\DeclareMathOperator{\M}{M}
\DeclareMathOperator{\SR}{SR}
\DeclareMathOperator{\sr}{sr}
\DeclareMathOperator{\shape}{shape}
\DeclareMathOperator{\Rad}{Rad}
\DeclareMathOperator{\Max}{Max}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Spin}{Spin}
\DeclareMathOperator{\Stab}{Stab}
\DeclareMathOperator{\ASR}{ASR}
\DeclareMathOperator{\asr}{asr}
\DeclareMathOperator{\Ums}{Ums}
\DeclareMathOperator{\Umd}{Umd}
\DeclareMathOperator{\rk}{rk}
\newcommand{\rA}{\mathsf{A}}
\newcommand{\rB}{\mathsf{B}}
\newcommand{\rC}{\mathsf{C}} 
\newcommand{\rD}{\mathsf{D}} 
\newcommand{\rE}{\mathsf{E}}

@@

We denote by $(\alpha, \beta)$ the scalar product of roots and by $\langle \alpha, \beta\rangle$ the integer $2(\alpha, \beta)/(\alpha, \alpha)$.
Denote by $W(\Phi)$ the subgroup of isometries of $\Phi$ generated by all reflections $\sigma_\alpha$ where $\sigma_\alpha(\beta)=\beta-\langle\alpha,\beta \rangle\cdot \alpha$.
For $S\subseteq \Phi$ denote by $W(S)$ the subgroup of $W(\Phi)$ generated by $\sigma_\alpha$ for $\alpha\in S$.

@@

\begin{rem} Let $\lambda_1, \lambda_2 \in \Lambda(\pi)$ be a pair of weights of a representation $\pi$ of a classical group such that $\lambda_1-\lambda_2\in \Phi$.
In this situation it will be convenient for us to write $x_{\lambda_1,\lambda_2}(\xi)$ instead of $x_{\lambda_1-\lambda_2}(\xi)$.
For example, for $\Phi=\rA_\ell$ we have $x_{1,2}(\xi)=x_{1-2}(\xi)=x_{\varpi_1 - \varpi_1 + \alpha_1}(\xi) = x_{\alpha_1}(\xi)$. \end{rem}

Let $I\leq R$ be an ideal. Denote by $\E(\Phi, I)$ the subgroup of $\G_{sc}(\Phi, R)$ generated by the elementary root unipotents $x_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in I$.
In the special case $I=R$ the subgroup $\E(\Phi, R)$ is called the {\it elementary subgroup} of the Chevalley group $\G_{sc}(\Phi, R)$.
Taddei's theorem asserts that $\E(\Phi, R)$ is a normal subgroup of $\G_{sc}(\Phi, R)$ provided $\Phi$ is an irreducible root system of rank $\geqslant 2$.

We denote by $\E(\Phi, R, I)$ the normal closure of $\E(\Phi, I)$ in $\E(\Phi, R)$, in other words $\E(\Phi, R, I) = \E(\Phi, I)^{\E(\Phi, R)}$.
Denote by $z_\alpha(s, \xi)$ the element $x_\alpha(s)^{x\ssub{-\alpha}(\xi)}$.

Since elementary subgroup $\E(\Phi,R,I)$ is normal in $\G(\Phi,R,I)$, one can define the $\K_1$-functor by $\K_1(\Phi,R,I)=\G(\Phi,R,I)/\E(\Phi,R,I)$. When $I=R$, we write $\K_1(\Phi,R)$ for $\K_1(\Phi,R,R)$. In some cases it is known that $\K_1(\Phi,R,I)$ is trivial. For example, $\SK_1(\ell+1,R)=\K(\rA_\ell,R)=1$ for any ring of stable rank $1$ (see Section~\ref{sec:stability-conditions}), and for other root systems some stronger assumption is required such as $\asr(R)=1$ or being semilocal. For a Euclidean ring $\K_1$ is trivial for every root system.

Let $k$ be a global field, $S$ be a set of places of $k$, $\mathcal{O}_S$ be the Dedekind ring of arithmetic type defined by $S$ and $I$ be an ideal of $\mathcal{O}_S$.
\begin{thm*}[{\cite[Theorem~3.6]{BassMilnorSerre} and \cite[Corollary~4.5]{Ma69}}]
Assume that the field $k$ has a real embedding. Then $\K_1(\Phi,\mathcal{O}_S,I)=1$.
\end{thm*}

%The following result of L.~Vaserstein and J.~Tits describes a set of generators of $\E(\Phi, R, I)$ (see e.\,g.~\cite[Theorem~2]{Va86}).
%\begin{thm}\label{theorem:Tits-Vaserstein} Let $\Phi$ be an irreducible root system of rank $\geq 2$, then the relative elementary subgroup $\E(\Phi, R, I)$
%is generated by the elements $z_\alpha(s, \xi)$, $\alpha\in \Phi$, $s\in I$, $\xi\in R$. \end{thm}

%In fact, one can choose a more economical set of generators for $\E(\Phi, R, I)$ (see~\cite[Theorem~3.4]{S}).
%\begin{thm}\label{theorem:Stepanov}
%Let $\Phi$ be a reduced irreducible root system of rank $\geq 2$ and let $S$ be any parabolic subset of roots of $\Phi$ with the special part $\Sigma_S$.
%Then $\E(\Phi, R, I)$ is generated by the union of the following two families of elements:
%\begin{itemize}
%\item $x_{\alpha}(s)$, where $s\in I$, $\alpha\in\Phi$;
%\item $z_\alpha(s,\xi)$, where $s\in I$, $\xi\in R$, $\alpha\in\Sigma_S$.
%\end{itemize}
%\end{thm}

For a special subset of roots $S\subseteq \Phi$ we denote by $\U(S, I)$ the subgroup spanned by all $x_{\alpha}(s)$ for $s\in I$, $\alpha\in S$.
The subgroup $\U(\Sigma_J, I)$ is normalized by $\E(\Delta_J, R)$, hence the Minkowski product set $\EP_J(R, I) = \E(\Delta_J, R, I) \cdot \U(\Sigma_J, I)$ is a subgroup. 
The following two equalities will be referred to in the sequel as {\it Levi decomposition}: 
\[ \EP_J(R, I) = \U(\Sigma_J, I) \cdot \E(\Delta_J, R, I) = \E(\Delta_J, R, I) \cdot \U(\Sigma_J, I). \]

@@

>>>>

%Math operators
\DeclareMathOperator{\K}{K}
\DeclareMathOperator{\SK}{SK}
\DeclareMathOperator{\G}{G}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\SL}{SL}
\DeclareMathOperator{\Sp}{Sp}
\DeclareMathOperator{\SO}{SO}
\DeclareMathOperator{\St}{St}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\EP}{EP}
\DeclareMathOperator{\Par}{P}
\DeclareMathOperator{\B}{B}
\DeclareMathOperator{\Hh}{H}
\DeclareMathOperator{\U}{U}
\DeclareMathOperator{\Z}{Z}
\DeclareMathOperator{\M}{M}
\DeclareMathOperator{\SR}{SR}
\DeclareMathOperator{\sr}{sr}
\DeclareMathOperator{\shape}{shape}
\DeclareMathOperator{\Rad}{Rad}
\DeclareMathOperator{\Max}{Max}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Spin}{Spin}
\DeclareMathOperator{\Stab}{Stab}
\DeclareMathOperator{\ASR}{ASR}
\DeclareMathOperator{\asr}{asr}
\DeclareMathOperator{\Ums}{Ums}
\DeclareMathOperator{\Umd}{Umd}
\DeclareMathOperator{\rk}{rk}
\newcommand{\rA}{\mathsf{A}}
\newcommand{\rB}{\mathsf{B}}
\newcommand{\rC}{\mathsf{C}} 
\newcommand{\rD}{\mathsf{D}} 
\newcommand{\rE}{\mathsf{E}}

@@

We denote by $(\alpha, \beta)$ the scalar product of roots and by $\langle \alpha, \beta\rangle$ the integer $2(\alpha, \beta)/(\beta, \beta)$.
Denote by $W(\Phi)$ the subgroup of isometries of $\Phi$ generated by all reflections $\sigma_\alpha$ where $\sigma_\alpha(\beta)=\beta-\langle\alpha,\beta \rangle\cdot \alpha$.
For $S\subseteq \Phi$ denote by $W(S)$ the subgroup of $W(\Phi)$ generated by $\sigma_\alpha$ for $\alpha\in S$.

@@

\begin{rem}
Let $\lambda_1, \lambda_2 \in \Lambda(\pi)$ be a pair of weights of a representation $\pi$ of a classical group such that $\lambda_1-\lambda_2\in \Phi$.
In this situation it will be convenient for us to write $x_{\lambda_1,\lambda_2}(\xi)$ instead of $x_{\lambda_1-\lambda_2}(\xi)$.
For example, for $\Phi=\rA_\ell$ we have $x_{1,2}(\xi)=x_{1-2}(\xi)=x_{\varpi_1 - \varpi_1 + \alpha_1}(\xi) = x_{\alpha_1}(\xi)$.
\end{rem}
For other classical root systems the elementary root unipotents are as follows.

In case $\Phi=\rC_\ell$ the long and short root unipotenrs are
\begin{align*}
& x_{i,-i}(\xi)=e+\xi e_{i,-i}, \\ 
& x_{i,j}(\xi)=e+\xi e_{ij}-\varepsilon_i\varepsilon_j\xi e_{-j,-i},\quad i\neq\pm j.
\end{align*}

Here $\varepsilon_i$ is the sign of $i$.

In case $\Phi=\rB_\ell,\rD_\ell$ the long and short root unipotents in $\SO(n,R)$ are
\begin{align*}
& x_{i,j}(\xi)=e+\xi e_{ij}-\xi e_{-j,-i},\quad i\neq\pm j,\ i,j\neq0, \\
& x_{i0}(\xi)=e+2\xi e_{i0}-\xi e_{0,i}-\xi^2 e_{i,-i},\quad i\neq0.
\end{align*}

\subsection{Relative Chevalley groups}\label{sec:relative-elementary}
Let $I\leq R$ be an ideal. Denote by $\E(\Phi, I)$ the subgroup of $\G_{sc}(\Phi, R)$ generated by the elementary root unipotents $x_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in I$.
In the special case $I=R$ the subgroup $\E(\Phi, R)$ is called the {\it elementary subgroup} of the Chevalley group $\G_{sc}(\Phi, R)$.

We denote by $\E(\Phi, R, I)$ the normal closure of $\E(\Phi, I)$ in $\E(\Phi, R)$, in other words $\E(\Phi, R, I) = \E(\Phi, I)^{\E(\Phi, R)}$.
Recall from the introduction that the elements $z_\alpha(s, \xi) = x_\alpha(s)^{x\ssub{-\alpha}(\xi)}$ generate $\E(\Phi, R, I)$ as an abstract group.

For a special subset of roots $S\subseteq \Phi$ we denote by $\U(S, I)$ the subgroup spanned by all $x_{\alpha}(s)$ for $s\in I$, $\alpha\in S$.
The subgroup $\U(\Sigma_J, I)$ is normalized by $\E(\Delta_J, R)$, hence the Minkowski product set $\EP_J(R, I) = \E(\Delta_J, R, I) \cdot \U(\Sigma_J, I)$ is a subgroup. 
The following two equalities will be referred to in the sequel as {\it Levi decomposition}: 
\begin{equation} \label{rel:Levi-decomp} \EP_J(R, I) = \U(\Sigma_J, I) \cdot \E(\Delta_J, R, I) = \E(\Delta_J, R, I) \cdot \U(\Sigma_J, I). \end{equation}
When $J = \{ \alpha_s \}$ for some $1 \leq s\leq \ell$ we use shorthand $\EP_s(R, I)$ for $\EP_{\{s\}}(R, I)$.
When $I=R$ we also write $\EP_J(R)$ instead of $\EP_J(R, R)$.

@@

\subsection{$\K_1$-functor modeled on Chevalley groups}

Recall that by Taddei's theorem $\E(\Phi, R)$ is a normal subgroup of $\G_{sc}(\Phi, R)$ provided $\Phi$ is an irreducible root system of rank $\geqslant 2$ (see~\cite{Ta}).

Using standard relativization argument it is not hard to deduce from Taddei's result that $\E(\Phi,R,I)$ is normal in $\G(\Phi,R,I)$ under the same assumptions on $\Phi$.
This allows us to define the relative $\K_1$-group as $\K_1(\Phi,R,I)=\G(\Phi,R,I)/\E(\Phi,R,I).$
When $I=R$, we write $\K_1(\Phi, R)$ for $\K_1(\Phi, R, R)$.

In some cases it is known that $\K_1(\Phi, R, I)$ is trivial.
For example, $\SK_1(\ell+1,R)=\K(\rA_\ell,R)=1$ for any ring of stable rank $1$ (see Section~\ref{sec:stability-conditions}), 
and for other root systems some stronger assumption is required such as $\asr(R)=1$ or being semilocal. 
For any Euclidean ring $\K_1(\Phi, R, I)$ is trivial for every root system $\Phi$.

Let $k$ be a global field, $S$ be a set of places of $k$, 
$\mathcal{O}_S$ be the Dedekind ring of arithmetic type defined by $S$ and $I$ be an ideal of $\mathcal{O}_S$.
\begin{prop} \label{prop:K1triv} Assume that the field $k$ has a real embedding. Then $\K_1(\Phi,\mathcal{O}_S,I)=1$. \end{prop}
\begin{proof} Follows from \cite[Theorem~3.6]{BassMilnorSerre} and \cite[Corollary~4.5]{Ma69}. \end{proof}
